---
title: "Salmon and Protected Areas"
author: "Adelaide Robinson"
date: "2022-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(here)
library(tidyverse)
library(sf)
library(tmap)
library(lfe)
```

# Background

# Data Wrangling: All Species

## Reading In Data

I used data from two sources: California Department of fish and wildlife and California Protected areas database.

```{r}
salmon_populations <- read_csv(here("data/Salmonid_Population_Monitoring_Data_CMPv2021.csv"))


watershed <- st_read(here("data/ds3001/ds3001.gdb"))


protected_areas <- st_read(here("data/CPAD_2022a/CPAD_2022a_Holdings.dbf"))

```

## Filtering to Adult Salmon Data

In this section I refine the data, for this study I am interested in looking at adult population counts, so I select data on Adults.

Not every watershed polygon provided by the CMP has count data for it. I am taking out polygons with no adult data associated. There are 157 polygons of watersheds, I have adult counts for 110 of them.

```{r}

spawning_data <- salmon_populations |> filter(`Life Stage` %in% "Adult") |> #all adult salmon
  select("Population", "Watershed", "Species", Brood_year = "Brood Year", "GEO_ID_POLY", "Value", "Metric", "Estimation method") |> #selecting relevant columns
  filter(!is.na(GEO_ID_POLY)) #taking out data with no matching spatial id

watershed_id <- unique(spawning_data$GEO_ID_POLY) #making a list of all the watersheds that have adult population data 

watershed_new <- watershed |> filter(GEO_ID_POL %in% watershed_id) |> st_make_valid() #filter to watersheds that have spawning data available


protected <- protected_areas |> select(UNIT_NAME, YR_EST) |> st_make_valid()
#selecting relevant columns

```

## Filtering Protected Areas

```{r}
protected2 <- protected |> filter(YR_EST < 1981| YR_EST %in% c(NA)) #remove after 1981

```

## Calculating Percent Protected: Geo spatial Wrangling

```{r}
#Filter to protected areas within watersheds of interest and find the area of intersection for each 
intersect_polygons <- st_intersection(protected2, watershed_new) |> 
   dplyr::select(Name, GEO_ID_POL) #select relevant columns
 
#find total area protected for each watershed 
total_overlap <- intersect_polygons |> group_by(Name) |> #group by watershed
  summarize(geometry = st_union(geometry))|> #combine geometries within watershed
  mutate(total_protected = st_area(geometry)) #find total protected 

# dropping geometry 
total_overlap_geomless <- total_overlap |> st_drop_geometry()

watershed_area <- watershed_new |> mutate(total_area = st_area(Shape)) #find the total area of each watershed 

watershed_protected <- left_join(watershed_area, total_overlap_geomless, by = "Name") #add area protected column by joining

#calculate percent protected 
watershed_final <- watershed_protected |> 
mutate(percent_protected = 
         as.numeric((total_protected)/ as.numeric(total_area)) *100) |> 
  mutate(percent_protected = round(percent_protected, digits = 0)) |> #round 
  mutate(percent_protected  = replace_na(percent_protected, 0)) #change NA to 0

#drop geometry and make it a data frame
watershed_geomless <- st_drop_geometry(watershed_final) |> as.data.frame() |> 
  select(- Method_Typ)

#combine spawning observations with percent protected
all_data <- left_join(spawning_data, watershed_geomless, by = c("GEO_ID_POLY" = "GEO_ID_POL")) |> select("Population", "Species", "Value", "percent_protected", "Name", "Brood_year", "Metric", "Estimation method")

```

```{r}
tmap_mode("view")

tmap_options(check.and.fix = TRUE)

tm_shape(total_overlap) + tm_fill(col = "#004600") + #map protected portions
  tm_shape(watershed_new) + tm_borders(col = "blue") + tm_add_legend(labels = c( "Watershed Boundary", "Protected area"), col = c("blue", "#004600")) #map watershed boundaries

```

# Steelhead

## Data Wrangling

```{r}
steelhead <- all_data |> filter(Species == "Steelhead") #filter for steelhead


steelhead_summary <- steelhead |>  group_by(Population, Brood_year) |> summarize(Value = max(Value), percent_protected = mean(percent_protected)) #taking the mean estimation for years where estimate done multiple ways 


data_years <- steelhead_summary |> group_by(Population) |> summarize(min_year = min(Brood_year), max_year = max(Brood_year), total_year = length(Brood_year)) #find what years I have data over


```

# 

## Regression

$populationcount = B_0 + B_1year_t + B_2Year_t * percentProtected +percentprotected +E_i$

```{r}

#running a regression with all the steelhead data
steelhead_all <- steelhead_summary |> mutate(year = as.numeric(Brood_year))

summary(lm(Value~year+ year:percent_protected + percent_protected, data = steelhead_all))

```

```{r}


model <- felm(Value~year+ year:percent_protected + percent_protected | Population, data = steelhead_all)

getfe(model)
```

## Assumptions of OLS

### Linear in Parameters

My data does not necessarily appear to be linear in parameters, which may cause it violate the first assumption of OLS.

```{r}
ggplot(data = steelhead_all, aes(x = year, y = Value)) + geom_point() + geom_smooth(method=lm, se = FALSE)

```

### X has variation

**This is true**

### Assumption 4

These do not appear to be normally distributed.

```{r}
model <- lm(Value~year + year:percent_protected, data = steelhead_final)
residuals <- model$residuals |> as.data.frame()


ggplot(data = residuals) + geom_histogram(aes(x = model$residuals))
```

# Coho

```{r}

coho <- all_data |> filter(Species == "Coho salmon")

coho2 <- coho |>  group_by(Population, Brood_year) |> summarize(Value = max(Value), percent_protected = max(percent_protected)) #taking the max estimation for years where estimate done in multiple ways 


data_years <- coho2 |> group_by(Population) |> summarize(min_year = min(Brood_year), max_year = max(Brood_year), total_year = length(Brood_year)) #find what years I have data over]


ggplot(data = coho_final) + geom_point(aes(x = Brood_year, y = Value))


```

```{r}
coho_final <- coho2 |>  mutate(year = as.numeric(Brood_year))

summary(felm(Value~year + year:percent_protected + percent_protected | Population, data = coho_final))


```

```{r}
acf(coho_final$Value, lag.max = 10)

```

# 

# Limitations and Issues

There were a lot of NA s in the year protected roughly 32 % of the polygons I used The definition of protected area was very loose. The data was not consistent over time, so I only used data from 2008-2018. Some watersheds within the study have hatchery releases which are not consistent over time, this can wildly affect the population from year to year.
