---
title: "Salmon and Protected Areas"
author: "Adelaide Robinson"
date: "2022-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(here)
library(tidyverse)
library(sf)
library(tmap)
library(kableExtra)
library(lfe)
```

# Background

# Data Wrangling: All Species

## Reading In Data

I used data from two sources: California Department of fish and wildlife and California Protected areas database.

```{r}
salmon_populations <- read_csv(here("data/Salmonid_Population_Monitoring_Data_CMPv2021.csv"))
#read in salmon population data 

watershed <- st_read(here("data/ds3001/ds3001.gdb"))
#read in watershed data
protected_areas <- st_read(here("data/CPAD_2022a/CPAD_2022a_Holdings.dbf"))
```

## Filtering to Adult Salmon Data

In this section I refine the data, for this study I am interested in looking at adult population counts, so I select data on Adults.

Not every watershed polygon provided by the CMP has count data for it. I am taking out polygons with no adult data associated. There are 157 polygons of watersheds, I have adult counts for 110 of them.

```{r}

spawning_data <- salmon_populations |> filter(`Life Stage` %in% "Adult") |> #all adult salmon
  select("Population", "Watershed", "Species", Brood_year = "Brood Year", "GEO_ID_POLY", "Value", "Metric", "Estimation method") |> #selecting relevant columns
  filter(!is.na(GEO_ID_POLY)) #taking out data with no matching spatial id

watershed_id <- unique(spawning_data$GEO_ID_POLY) #making a list of all the watersheds that have adult population data 

watershed_new <- watershed |> filter(GEO_ID_POL %in% watershed_id) |> st_make_valid() #filter to watersheds that have spawning data available


protected <- protected_areas |> select(UNIT_NAME, YR_EST)
#selecting relevant columns

```

## Filtering Protected Areas

```{r}
protected2 <- protected |> filter(YR_EST < 1981| YR_EST %in% c(NA)) #remove after 1981

```

## Calculating Percent Protected: Geo spatial Wrangling

```{r}
#Filter to protected areas within watersheds of interest and find the area of intersection for each 
intersect_polygons <- st_intersection(watershed_new, protected2) |> 
   dplyr::select(Name, GEO_ID_POL) #select relevant columns
 
#find total area protected for each watershed 
total_overlap <- intersect_polygons |> group_by(Name) |> #group by watershed
  summarize(geometry = st_union(Shape))|> #combine geometries within watershed
  mutate(total_protected = st_area(geometry)) #find total protected 

# dropping geometry 
total_overlap_geomless <- total_overlap |> st_drop_geometry()

watershed_area <- watershed_new |> mutate(total_area = st_area(Shape)) #find the total area of each watershed 

watershed_protected <- left_join(watershed_area, total_overlap_geomless, by = "Name") #add area protected column by joining

#calculate percent protected 
watershed_final <- watershed_protected |> 
mutate(percent_protected = 
         as.numeric((total_protected)/ as.numeric(total_area)) *100) |> 
  mutate(percent_protected = round(percent_protected, digits = 0)) |> #round 
  mutate(percent_protected  = replace_na(percent_protected, 0)) #change NA to 0


#drop geometry and make it a data frame
watershed_geomless <- st_drop_geometry(watershed_final) |> as.data.frame() |> 
  select(- Method_Typ)

#combine spawning observations with percent protected
all_data <- left_join(spawning_data, watershed_geomless, by = c("GEO_ID_POLY" = "GEO_ID_POL")) |> select("Population", "Species", "Value", "percent_protected", "Name", "Brood_year", "Metric", "Estimation method", "GEO_ID_POLY", "total_area")

```

```{r}


tmap_mode("view")

tmap_options(check.and.fix = TRUE)

tm_shape(total_overlap) + tm_fill(col = "#004600") + #map protected portions
  tm_shape(watershed_new) + tm_borders(col = "blue") + tm_add_legend(labels = c( "Watershed Boundary", "Protected area"), col = c("blue", "#004600")) #map watershed boundaries

```

```{r}
tm_shape(watershed_final) + tm_polygons(col = "percent_protected")
```

# Steelhead

## Data Wrangling

```{r}
steelhead <- all_data |> filter(Species == "Steelhead") #filter for steelhead

steelhead_consistent <- steelhead |> group_by(Population) |> summarize(percent_protected = max(percent_protected), area = max(total_area)) #making value of percent protected consistent across the population

steelhead_summary <- steelhead |>  group_by(Population, Brood_year) |> summarize(Value = round(mean(Value),0)) #taking average where population count calculated in more than one way

steelhead_final <- left_join(steelhead_summary, steelhead_consistent, by = "Population")
        
No_fish <- steelhead_final |> group_by(Population) |> summarize(max = max(Value)) |> filter(max > 0)

Keep <- No_fish$Population

steelhead <- steelhead_final |>  filter(Population %in% Keep)
                                                    
```

## option including all years

```{r}

steelhead_all <- steelhead_final |>  mutate(year = as.numeric(Brood_year) - 1981)

```

## Regression

$populationcount = B_0 + B_1year_t + + B_2percentprotected + B_3Year_t * percentProtected +E_i$

```{r}

model <- lm(Value~percent_protected + year + year:percent_protected, data = steelhead_all)

tidy(model, conf.int = TRUE)

model <- lm(Value~area + year + percent_protected + year:percent_protected, data = steelhead_all)

tidy(felm(Value~percent_protected:year + year | population, data = steelhead_all))
```

# all data regressions

```{r}
model <- felm(Value~ year + year:percent_protected | Population, data = steelhead_all)
summary(model)



model <- felm(Value~percent_protected + area |year |0|year, data = steelhead_all)

tidy(model)
```

## Assumptions of OLS

### Linear in Parameters

My data does not necessarily appear to be linear in parameters, which may cause it violate the first assumption of OLS.

```{r}
ggplot(data = steelhead_all, aes(x = year, y = Value)) + geom_point() +labs(x = "Year", y = "Population Count") + ylim(0,5000)
```

### X has variation

**This is true**

### Assumption 4

These do not appear to be normally distributed.

```{r}

residuals <- model$residuals |> as.data.frame()


ggplot(data = residuals) + geom_histogram(aes(x = model$residuals)) + labs(x = "residuals")
```

```{r}

res = resid(model)

ggplot() + geom_point(aes(x = steelhead_all$percent_protected,
                            y = res))


```


```{r}
lm_model <- lm(Value~year + percent_protected + year:percent_protected, data = steelhead_all)
summary(lm_model)

felm_protected <- felm(Value~percent_protected |year|0|year, data = steelhead_all)
tidy(felm_protected, conf.int = TRUE)

time_model <- felm(Value~year|Population, data = steelhead_all)

tidy(time_model)

percent_model <- felm(Value~year + year:percent_protected |Population|0|Population, data = steelhead_all)

summary(percent_model)
```

# Coho

```{r}

coho <- all_data |> filter(Species == "Coho salmon")

coho_consistent <- steelhead |> group_by(Population) |> summarize(percent_protected = max(percent_protected), area = max(area)) #making value of percent protected consistent across the population

coho_summary <- coho |>  group_by(Population, Brood_year) |> summarize(Value = round(mean(Value),0)) #taking average where population count calculated in more than one way

coho_final <- left_join(coho_summary, coho_consistent, by = "Population")


coho12 <- coho_summary |> group_by(Population) |> summarize(max = max(Value)) |> filter(max == 0)

remove <- coho12$Population

coho_usable <- coho_final |> filter(!Population %in% remove)

coho_final <- coho_usable |>  mutate(year = as.numeric(Brood_year) - 1981)

```

```{r}


model <- lm(Value~year + percent_protected + year:percent_protected, data = coho_final)


model <- felm(Value~year + year:percent_protected |Population, data = coho_final)

summary(lm(Value~area, data = coho_final))
```

## all data 

```{r}
coho_all <- coho2 |>  mutate(year = as.numeric(Brood_year) - 1981)
```

# Limitations and Issues

There were a lot of NA s in the year protected roughly 32 % of the polygons I used The definition of protected area was very loose. The data was not consistent over time, so I only used data from 2008-2018. Some watersheds within the study have hatchery releases which are not consistent over time, this can wildly affect the population from year to year.
